---
title: "Production Validation: {Description}"
date: YYYY-MM-DD
categories: [production-validation]
status: pass          # pass | fail | partial-pass | in-progress
branch: "main"        # branch or tag of the pipeline that generated the data
---

<!--
TEMPLATE — copy to testing-log/{date}-{description}.qmd.
Use for scientific comparisons: new pipeline vs legacy workflow, a prior production
run, or any external ground truth. Fill in the parameters, findings, and optionally
include analysis code with eval: false.

For entries with executable R analysis (plots, tables), either:
  (a) Set eval: false globally and embed key results as static text, OR
  (b) Place data files in docs/testing-log/data/ and run quarto render to freeze.

See docs/testing-workflow.qmd for instructions.
-->

One paragraph summarising what was compared, the mapping panel and genetic architectures
tested, and the overall conclusion.

**Conclusion:** [State the outcome in one sentence.]

---

## Summary

| Metric | Value |
|--------|-------|
| Median absolute Power difference | — |
| Median absolute FDR difference | — |
| Max absolute Power difference | — |
| Max absolute FDR difference | — |
| Between-workflow vs within-workflow variation | — |

---

## Parameters

| Parameter | Value |
|-----------|-------|
| Species / release | — |
| Mapping panel size | — |
| nQTL range | — |
| h² values | — |
| Replicates (new pipeline) | — |
| Replicates (reference) | — |
| Effect distribution | — |
| MAF | — |
| Mapping method compared | — |
| Significance threshold | — |

Pipeline run location (if HPC):

---

## Reference Data

| Dataset | Source | Description |
|---------|--------|-------------|
| New pipeline | `db_simulation_assessment_results.tsv` | Output from nemascan-sims-nf production run |
| Reference | — | Legacy workflow / prior run / external ground truth |

::: {.callout-note}
Data files are not committed to this repository.
To reproduce the analysis, place the required files in `docs/testing-log/data/raw/` and re-render.
:::

---

## Statistical Design

Describe any downsampling, bootstrapping, or matched-replicate approach used to make the comparison fair (e.g., correcting for replicate imbalance between workflows).

---

## Results

### Key metrics

| Metric | Value |
|--------|-------|
| — | — |

### Interpretation

Narrative summary of the results. Are the workflows concordant? What is the range of discordance across architectures?

---

## Analysis Code

```{r}
#| label: setup
#| eval: false

library(data.table)
library(dplyr)
library(ggplot2)

# Update paths to match your data location
raw_dir <- here::here("docs", "testing-log", "data", "raw")
new_pipeline_file <- file.path(raw_dir, "db_simulation_assessment_results.tsv")
reference_file    <- file.path(raw_dir, "reference_assessment.tsv")
```

```{r}
#| label: load-data
#| eval: false

new_df <- data.table::fread(new_pipeline_file)
ref_df <- data.table::fread(reference_file)
```

```{r}
#| label: comparison
#| eval: false

# Add analysis code here
```

---

## Notes

<!-- Any caveats, known limitations, or follow-up items. -->
