
```{r}
library(data.table)
library(dplyr)
```
Code to check the mapping data that is input to the Assess_Sims.R script to troubleshoot the issue with the var.exp() function and NA values.
Error message:
```
  Error in lm.fit(x, y, offset = offset, singular.ok = singular.ok, ...) : 
    0 (non-NA) cases
  Calls: %>% ... summary -> aov -> eval -> eval -> <Anonymous> -> lm.fit
  Execution halted
```

# Load the simulation data 
Load the effects file (.par) file created by GCTA 
```{r}
effects_file <- "20250527_Troubleshoot_Assess/5_2_0.2_0.05_gamma_cb.strain.set.290_literal.harborporpoise_sims.par"

effects <- fread(effects_file, header = TRUE)
head(effects)
```
Load the phenotype file (.pheno) file created by GCTA
```{r}
pheno_file <- "20250527_Troubleshoot_Assess/5_2_0.2_0.05_gamma_cb.strain.set.290_literal.harborporpoise_sims.pheno"

phenos <- data.table::fread(
  pheno_file,
  header = FALSE,
  col.names = c("strain", "strain2", "trait.value")
) %>%
  dplyr::select(-strain2)
```
Load the genotype matrix
```{r}
gm_file <- "20250527_Troubleshoot_Assess/cb.strain.set.290_literal.harborporpoise_0.05_Genotype_Matrix.tsv"

gm <- data.table::fread(gm_file)
```
# Create the complete effects data frame
The complete effects df is generated by combining the genomtype matrix with the .par and .pheno files
```{r}
complete.effects <- gm %>%
  tidyr::unite("QTL", c(CHROM, POS), sep = ":", remove = F) %>%
  dplyr::filter(QTL %in% effects$QTL) %>% # Filter markers in the GM that are causal variants
  dplyr::select(-CHROM, -POS) %>% # Get rid of the CHOM and POS columns
  tidyr::pivot_longer(cols = !c(QTL, REF, ALT), names_to = "strain", values_to = "allele") %>%
  dplyr::full_join(., effects) %>%
  dplyr::full_join(., phenos) %>%
  dplyr::group_by(QTL) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::filter(
    n != 1,
    !is.na(QTL)
  )
head(complete.effects)
```
# Create the genos.effects dataframe
```{r}
genos.effects <- gm %>%
  tidyr::unite("QTL", c(CHROM, POS), sep = ":", remove = F) %>%
  dplyr::filter(QTL %in% effects$QTL) %>%
  dplyr::select(-CHROM, -POS) %>%
  tidyr::pivot_longer(cols = !c(QTL, REF, ALT), names_to = "strain", values_to = "allele") %>%
  dplyr::full_join(., effects) %>%
  dplyr::full_join(., phenos) %>%
  dplyr::group_by(QTL) %>%
  tidyr::nest()
```
# Define the `var.exp` function
```{r}
# Simulated Variance Explained
var.exp <- function(data, QTL) {
  aov.out <- summary(aov(data = data, trait.value ~ allele + 1))
  SST <- sum(aov.out[[1]][[2]])
  SSallele <- aov.out[[1]][[2]][[1]]
  simulated.variance.exp <- SSallele / SST
  data.frame(QTL, simulated.variance.exp)
}
```
# Apply the function with `purr`
```{r}
# Calculate the variance explained for each QTL based on the strains and phenotyes
simQTL.variance.explained <- purrr::map2(
  genos.effects$data,
  genos.effects$QTL,
  var.exp
) %>%
  Reduce(rbind, .)
```
# Debugging the `var.exp` function
The issue is with index 3 in the data, so lets inspect the third element of genos.effects data.
## genos.effects elements
```{r}
genos.effects$data[[3]]
```
```{r}
genos.effects$data[[1]]
```
Okay lots of differences here between the working entry compared to the entry we get an error with. Mainly, there is no data for the third entry `3:948637`

This is probably an error with steps to generate the genos.effects dataframe.
## Troubleshooting creation of genos.effcts
First we perfrom some basic processing of the genotype matrix by combining the CHROM and pos and filtering to just the simulated causal variants
```{r}
ge1 <- gm %>%
  tidyr::unite("QTL", c(CHROM, POS), sep = ":", remove = F) %>%
  dplyr::filter(QTL %in% effects$QTL)
nrow(ge1)
```
Interesting, there are only 2 SNPs that are retained in the genotype matrix even though there are 5 total SNPs in the `genos.effects` file. 
