```{r}
library(data.table)
library(dplyr)
```

# Checking the intermediate file for the process with the issue
Wordir for the simulation rep that threw the error as recorded in the .nextflow log: `/scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/8d/cda5b6ff0e1d9a2b4016bc21599ad6`

```{r}
```


# Checking the files generated by `PLINK_RECODE_VCF` for the mapping panel
## Inputs
Paths to the plink files generated for the mapping panel with the issue
```{r}
to_sims_bed_file <- "20250527_Troubleshoot_Assess/PLINK_RECODE_VCF/TO_SIMS.bed"
to_sims_bim_file <- "20250527_Troubleshoot_Assess/PLINK_RECODE_VCF/TO_SIMS.bim"
markers_file <- "20250527_Troubleshoot_Assess/PLINK_RECODE_VCF/markers.txt"
plink_prune_in_file <- "20250527_Troubleshoot_Assess/PLINK_RECODE_VCF/plink.prune.in"

effects_file <- "20250527_Troubleshoot_Assess/R_ASSESS_SIMS/5_2_0.2_0.05_gamma_cb.strain.set.290_literal.harborporpoise_sims.par"
```
Causal variant missing from the genotype matrix
```{r}
effects <- data.table::fread(effects_file, header = TRUE)
cvs <- effects %>%
  dplyr::pull(QTL)
```

## Check `markers.txt` file
This is the file passed to the process used to create the genotype matrix so we would expect that the variant would be absent here
```{r}
markers <- data.table::fread(
  markers_file,
  header = FALSE,
  col.names = c("chrom", "pos")
) %>%
  # combine the chromosome and position into a single column
  dplyr::mutate(
    QTL = paste(chrom, pos, sep = ":")
  )

markers %>%
  dplyr::filter(QTL %in% cvs)
```
Three of the five causal variants are missing from the markers.txt file.

Lets check if they are present in the plink.prune.in file used to generate the markers.txt file
## Check `plink.prune.in`
```{r}
plink_prune_in <- data.table::fread(
  plink_prune_in_file,
  header = FALSE,
  col.names = c("QTL")
)
```
Check if the causal variants are present in the plink.prune.in file
```{r}
plink_prune_in %>%
  dplyr::filter(QTL %in% cvs)
```
The same three causal variants are missing from the plink.prune.in file.
## Check `TO_SIMS.bim`
This is the file used to select the causal variants for the mapping panel so we would expect that all the causal variants are present here
```{r}
to_sims_bim <- data.table::fread(
  to_sims_bim_file,
  header = FALSE,
  col.names = c("chrom", "marker", "x", "pos", "ref", "alt")
)

to_sims_bim %>%
  dplyr::filter(marker %in% cvs)
```
Interesting, we would expect that the `TO_SIMS.bim` file would contain all the causal variants but it only contains two of the five. 

Likely there is an issue with the code to select the causal variants for this mapping panel. Either the code is not selecting the correct variants or the variants are present in the input file passed to the script to select the causal variants.

# Checking the input files to `PYTHON_SIMULATE_EFFECTS_GLOBAL`
Workdir for the simulation rep that threw the error: 
From the NF log file for the run
```
May-27 13:17:34.136 [Task submitter] DEBUG nextflow.executor.GridTaskHandler - [SLURM] submitted process PYTHON_SIMULATE_EFFECTS_GLOBAL (5 2 gammacb.strain.set.290_literal.harborporpoise_0.05) > jobId: 3127370; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/de/c52cf5c018104428c660cd730f59b1
****
```
Okay this is really strange - the causal variants that are selected by the process are different from the causal variants that are included in the `.par` file that is passed to the assess sims process.

# Create a log file for the simulation rep that threw the error
```{bash}
cat .nextflow.log | grep -e cb.strain.set.290_literal.harborporpoise > error_popid.nextflow.log
```

# Checking the intermediate files for `GCTA_SIMULATE_PHENOTYPES`
Workdir for the simulation rep that threw the error:
From the NF log file for the run
```
May-27 13:17:47.217 [Task monitor] DEBUG n.processor.TaskPollingMonitor - Task completed > TaskHandler[jobId: 3127406; id: 153; name: GCTA_SIMULATE_PHENOTYPES (5 2 0.2 gamma cb.strain.set.290_literal.harborporpoise_0.05); status: COMPLETED; exit: 0; error: -; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/5e/1ea80dee95efceea22e1596577d9bf started: 1748366267118; exited: 2025-05-27T17:17:45.881468852Z; ]
```
Here the .par file and the causal variants are the same. Maybe this issue is occuring in a latter process and the input channels are getting mixed up somewhere?

# Checking the relevant intermediate files for `PLINK_UPDATE_BY_H2` 
Workdir for the simulation rep that threw the error:
From the NF log file for the run
```
May-27 13:17:51.637 [Task submitter] DEBUG nextflow.executor.GridTaskHandler - [SLURM] submitted process PLINK_UPDATE_BY_H2 (5 2 0.2 gamma cb.strain.set.290_literal.harborporpoise_0.05) > jobId: 3127465; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/70/c741c6b337a5dc1e4b92aef89b05f1
```

# Checking the relevant intermediate files for `GCTA_MAKE_GRM`
```
May-27 13:18:02.958 [Task submitter] DEBUG nextflow.executor.GridTaskHandler - [SLURM] submitted process GCTA_MAKE_GRM (5 2 0.2 gamma inbred cb.strain.set.290_literal.harborporpoise_0.05) > jobId: 3127526; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/5d/a4e1b7c213b98ed8312b1e730d7bb4
```
# Checking the relevant intermediate files for `PYTHON_CHECK_VP`
```
May-27 13:18:12.331 [Task submitter] DEBUG nextflow.executor.GridTaskHandler - [SLURM] submitted process PYTHON_CHECK_VP (5 2 0.2 gamma inbred cb.strain.set.290_literal.harborporpoise_0.05) > jobId: 3127609; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/4c/957d44a05a91823a5e46d6114d95d1
```

# Checking the relevant intermediate files for `GCTA_PERFORM_GWA`
```
May-27 13:18:22.035 [Task submitter] DEBUG nextflow.executor.GridTaskHandler - [SLURM] submitted process GCTA_PERFORM_GWA (5 2 0.2 gamma inbred pca cb.strain.set.290_literal.harborporpoise_0.05) > jobId: 3127665; workDir: /scratch4/eande106/Ryan/2025_briggsae_gwas_strainset_sims/bc/93ab77bc1c8d5fbbcaadb0cde19121
```
