---
title: "Functional Validation: Non-Marker Causal Variants"
subtitle: "feature/add-nonmarker-cvs → feature-database-integration"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: false
execute:
  eval: false
  echo: true
---

## 1. Overview

The `feature/add-nonmarker-cvs` branch extends the pipeline to allow causal variants
(CVs) to be drawn from a separately-configured variant pool that is independent of the
GWA marker SNP set. Previously all causal variants had to be present in the marker SNP
set — they were the same PLINK binary. This restriction was lifted by introducing a
second, parallel PLINK run (the **CV pool**) for each strain group, whose MAF and LD
thresholds can be set independently of the GWA marker set.

This branch is **22 commits ahead** of `feature-database-integration` (13 step-branch
merges, fixup commits, and new test data). Changes touch: the strainfile format, CLI
parameters, the PLINK module, `main.nf` channel routing, Python effect simulation, GCTA
phenotype inputs, the R hash framework (`database.R`), three DB-write R scripts and
their Nextflow modules, and the test suite.

This notebook covers:

1. Strainfile format (2-column → 6-column)
2. Marker set ID hash — v=2 function demonstration (live R evaluation)
3. Dual PLINK pipeline wiring
4. CV binary routing through simulation
5. Unit test results
6. Pipeline stub runs (single-species `test` and three-species `test_three_species`)

---

## 2. Strainfile Format

### Before (2-column, `feature-database-integration`)

The old format was a space-separated file with no header. Each line had a group name and
a comma-separated strain list. VCF source, species, MAF threshold, and LD threshold were
all supplied via CLI parameters (`--vcf`, `--species`, `--maf`, `--ms_ld`).

```
ce.test.200strains AB1,BRC20067,BRC20263,...
```

### After (6-column TSV, this branch)

The new format is a tab-separated file **with a header row**. Each row is a complete,
self-contained strain group specification. The `--vcf`, `--species`, `--maf`,
`--ms_maf`, and `--ms_ld` CLI parameters are removed.

```
group	species	vcf	ms_maf	ms_ld	strains
ce.test.200strains	c_elegans	data/test/test.20210901.vcf.gz	0.05	0.8	AB1,BRC20067,...
```

Column definitions:

| Column | Type | Description |
|--------|------|-------------|
| `group` | string | Unique strain group identifier; used in file naming and DB keys |
| `species` | string | `c_elegans`, `c_briggsae`, or `c_tropicalis`; controls CaeNDR URL construction |
| `vcf` | string | 8-digit CaeNDR release date **or** path to a VCF file |
| `ms_maf` | float | Minor allele frequency threshold for the **marker SNP set** |
| `ms_ld` | float | LD R² pruning threshold for the **marker SNP set** |
| `strains` | string | Comma-separated strain list |

### Three-species strainfile

The new `test_three_species` profile uses `data/test/test_strains_three_species.txt`
with 9 groups spanning all three supported *Caenorhabditis* species:

```
group	species	vcf	ms_maf	ms_ld	strains
ce.test	c_elegans	data/test/test_ce.20250625.vcf.gz	0.05	0.8	AB1,BBA2,...
ce.test.popA	c_elegans	data/test/test_ce.20250625.vcf.gz	0.05	0.8	AB1,BBA2,...
ce.test.popB	c_elegans	data/test/test_ce.20250625.vcf.gz	0.05	0.8	ECA1937,...
cb.test	c_briggsae	data/test/test_cb.20250626.vcf.gz	0.05	0.8	AF16,BRC20069,...
cb.test.popA	c_briggsae	data/test/test_cb.20250626.vcf.gz	0.05	0.8	AF16,...
cb.test.popB	c_briggsae	data/test/test_cb.20250626.vcf.gz	0.05	0.8	ECA1143,...
ct.test	c_tropicalis	data/test/test_ct.20250627.vcf.gz	0.05	0.8	BRC20391,...
ct.test.popA	c_tropicalis	data/test/test_ct.20250627.vcf.gz	0.05	0.8	BRC20391,...
ct.test.popB	c_tropicalis	data/test/test_ct.20250627.vcf.gz	0.05	0.8	JU2505,...
```

Because `species` and `vcf` are per-row, a single pipeline invocation can mix
*C. elegans*, *C. briggsae*, and *C. tropicalis* groups.

---

## 3. Marker Set ID — v=2 Hash Function

The marker set ID hash was upgraded from v=1 (2 parameters: `population`, `maf`) to
v=2 (5 parameters: `population`, `maf`, `species`, `vcf_release_id`, `ms_ld`). The
version prefix is encoded in the hash input string, making v=1 and v=2 hashes
structurally incompatible — any existing DB built with v=1 hashes must be regenerated.

```{r}
#| eval: false
Sys.setenv(NEMASCAN_R_LIB = normalizePath("../R"))
source("../R/setup.R")

# v=2: now includes species, vcf_release_id, ms_ld
ms <- generate_marker_set_id("ce.test", 0.05, "c_elegans", "20220216", 0.8)
ms$hash         # → "db39e634ef0bfa76e1bd"
ms$hash_string  # → "v=2|population=ce.test|maf=0.0500000000|..."
```

```{r}
#| eval: false
# Show trait and mapping ID chain
tr  <- generate_trait_id(ms$hash, 5, "gamma", 1, 0.8)
map <- generate_mapping_id(tr$hash, "inbred", TRUE)

cat("marker_set_id :", ms$hash, "\n")
cat("trait_id      :", tr$hash, "\n")
cat("mapping_id    :", map$hash, "\n")
```

```{r}
#| eval: false
# Sensitivity: changing species changes the marker set hash
ms_cb <- generate_marker_set_id("ce.test", 0.05, "c_briggsae", "20220216", 0.8)
cat("c_elegans hash :", ms$hash, "\n")
cat("c_briggsae hash:", ms_cb$hash, "\n")
cat("hashes equal?  :", ms$hash == ms_cb$hash, "\n")  # → FALSE
```

```{r}
#| eval: false
# CV parameters are NOT in the hash — runs differing only in cv_maf/cv_ld reuse the same marker set record
ms2 <- generate_marker_set_id("ce.test", 0.05, "c_elegans", "20220216", 0.8)
cat("same hash on rerun?:", ms$hash == ms2$hash, "\n")  # → TRUE
```

::: {.callout-important}
**CV parameters are not included in the marker set ID.** Runs differing only in
`--cv_maf` / `--cv_ld` reuse the same marker SNP set record. This is intentional —
the GWA marker panel is independent of which variants are drawn as causal.
:::

---

## 4. Dual PLINK Pipeline

### Aliased module imports

`main.nf` imports `PLINK_RECODE_VCF` twice under different aliases:

```groovy
include { PLINK_RECODE_VCF as PLINK_RECODE_MS_VCF } from './modules/plink/recode_vcf/main'
include { PLINK_RECODE_VCF as PLINK_RECODE_CV_VCF } from './modules/plink/recode_vcf/main'
```

### Channel fan-out

After VCF renaming, the renamed VCF is forked with `.tap{}` before the two parallel
PLINK calls to avoid queue-channel double-consumption:

```groovy
BCFTOOLS_RENAME_CHROMS.out.vcf
    .tap { ch_renamed_for_ms }
    .tap { ch_renamed_for_cv }

// Marker SNP set — uses per-group ms_maf and ms_ld from strainfile
PLINK_RECODE_MS_VCF(
    ch_renamed_for_ms,
    ch_mito_num,
    ch_sf.ms_maf_vals.map { meta, ms_maf -> ms_maf },
    ch_sf.ms_ld_vals.map  { meta, ms_ld  -> ms_ld  }
)

// Causal variant pool — uses global cv_maf/cv_ld (or per-group ms_maf fallback)
PLINK_RECODE_CV_VCF(
    ch_renamed_for_cv,
    ch_mito_num,
    ch_sf.cv_maf_vals.map { meta, cv_maf_eff -> cv_maf_eff },
    Channel.value(cv_ld)
)
```

### Module change: `each maf` → `val maf`

The PLINK module input was changed from `each maf` (which creates a Cartesian product
over a list of MAF values) to `val maf` (which receives a single pre-resolved value per
upstream item). A new `val ld` input was added for the LD threshold:

```groovy
// modules/plink/recode_vcf/main.nf
input:
tuple val(meta), path(vcf), path(vcf_index)
val mito_name
val maf          // was: each maf — val runs once per upstream item (no Cartesian product)
val ld           // LD R² threshold for --indep-pairwise; was hardcoded 0.8
```

### Data-flow diagram

```
VCF (per group)
  │
  ▼
BCFTOOLS_EXTRACT_STRAINS → BCFTOOLS_RENAME_CHROMS
                                        │
                    ┌───────────────────┤ .tap{}
                    ▼                   ▼
           ch_renamed_for_ms   ch_renamed_for_cv
                    │                   │
                    ▼                   ▼
        PLINK_RECODE_MS_VCF   PLINK_RECODE_CV_VCF
        (ms_maf, ms_ld)        (cv_maf, cv_ld)
                    │                   │
              TO_SIMS.*           CV_TO_SIMS.*
        (GWA marker set)        (CV variant pool)
```

### Process selector (`nextflow.config`)

`ext.tag` disambiguates SLURM job names when both aliases run for the same group:

```groovy
process {
    withName: 'PLINK_RECODE_MS_VCF' {
        ext.tag = { "${meta.id}_MS" }
    }
    withName: 'PLINK_RECODE_CV_VCF' {
        ext.tag = { "${meta.id}_CV" }
    }
}
```

---

## 5. CV Binary Routing Through Simulation

### Why GCTA must use the CV binary

Causal variants are sampled from `CV_TO_SIMS.bim` by `PYTHON_SIMULATE_EFFECTS_GLOBAL`.
These variants may **not** be in `TO_SIMS.bim` (the marker SNP set) if `cv_maf < ms_maf`
or if LD pruning removed them from the marker set. Because GCTA's `--simu-qt` requires
the causal variants to be present in the input `--bfile`, phenotype simulation must use
the CV binary (`--bfile CV_TO_SIMS`), not the marker set binary.

### `.tap{}` fix for three-subscriber queue channel

`ch_sf.marker_set_params` is a Nextflow queue channel emitting one item per strain
group. In DSL2, queue channels are consumed once — if three downstream processes each
try to read from the same channel, emissions are distributed round-robin instead of
broadcast. The fix uses `.tap{}` to clone the channel before any subscriber reads it:

```groovy
// ch_sf.marker_set_params is a queue channel — fan out with .tap{} before
// 3 subscribers consume it (write_marker_set, write_genotype_matrix, write_gwa_to_db).
// Without this, DSL2 distributes emissions round-robin among competing readers.
ch_sf.marker_set_params
    .tap { ch_marker_set_params_for_ms }
    .tap { ch_marker_set_params_for_gm }
    .set { ch_marker_set_params_for_gwa }
```

Each of the three DB-write processes receives a complete, independent copy of the
marker-set parameter tuples.

---

## 6. Unit Tests

```bash
Rscript tests/run_tests.R
```

**Pre-captured output (2026-03-03):**

```
assessment_cross_validation: SSSSS
assessment: .....................................................................
cross_validation: SSSSSS
db_structure: SSSSSSSSSSSSSSSSSSS
genotype_storage: ...................
hash_generation: ...........................................................................
library_check: .................
module_sourcing: ...............
read_bim: ...
read_raw_gwa: ..........
safe_log10p: .........
schema_changes: .....................
significance: ...
trait_storage: ...........................................
write_gwa_to_db: ......

══ Skipped ═════════════════════════════════════════════════════════════════════
  (30 integration tests skipped — require TEST_DB_DIR / TEST_LEGACY_ASSESSMENT env vars)

══ DONE ════════════════════════════════════════════════════════════════════════
```

- `.` = pass, `S` = skipped (integration test, requires env vars)
- `hash_generation` (71 tests) includes golden-value assertions:
  - `generate_marker_set_id("ce.test", 0.05, "c_elegans", "20220216", 0.8)$hash` → `"db39e634ef0bfa76e1bd"`
  - `generate_trait_id("db39e634ef0bfa76e1bd", 5, "gamma", 1, 0.8)$hash` → `"aadfdfe371a917896ad6"`
  - `generate_mapping_id("aadfdfe371a917896ad6", "inbred", TRUE)$hash` → `"c7e06e757cd1102e6092"`
- The three test files updated in this PR: `test-genotype_storage.R`, `test-trait_storage.R`,
  and `test-write_gwa_to_db.R` were updated to use the 5-parameter v=2 signature and new
  golden values.

---

## 7. Pipeline Stub Runs

### 7a. Single-species (`test` profile)

```bash
nextflow run . -profile test -stub-run
```

> Note: Nextflow `>=24.10.0,<25.0.0` is required (manifest guard). The `>>` publish
> block syntax is a 24.x preview feature not supported in 25.x.

**Pre-captured output (2026-03-03):**

```
N E X T F L O W  ~  version 24.10.5
Launching `./main.nf` DSL2 - revision: 85f6398865

Strainfile                              = .../data/test/test_strains.txt
Number of QTLs/phenotype simulated      = .../data/test/fixed_architecture/nqtl.csv
Phenotype heritability file             = .../data/test/fixed_architecture/h2.csv
Number of replicates to simulate        = 1
Causal variant MAF                      = (per-group ms_maf)
Causal variant LD threshold             = 0.8

executor >  local (48)

BCFTOOLS_CREATE_GENOTYPE_MATRIX (ce.test.200strains_0.05)           ✔
BCFTOOLS_EXTRACT_STRAINS (ce.test.200strains)                        ✔
BCFTOOLS_RENAME_CHROMS (ce.test.200strains)                          ✔
DB_MIGRATION_AGGREGATE_METADATA (aggregate)                          ✔
DB_MIGRATION_ANALYZE_QTL × 8   (BF/EIGEN × inbred/loco × pca/nopca) ✔
DB_MIGRATION_ASSESS_SIMS × 8                                         ✔
DB_MIGRATION_WRITE_GENOTYPE_MATRIX (ce.test.200strains_0.05)         ✔
DB_MIGRATION_WRITE_GWA_TO_DB × 4  (inbred/loco × pca/nopca)         ✔
DB_MIGRATION_WRITE_MARKER_SET (ce.test.200strains_0.05)              ✔
DB_MIGRATION_WRITE_TRAIT_DATA (ce.test.200strains_5_1_0.8)           ✔
GCTA_MAKE_GRM × 2              (inbred, loco)                        ✔
GCTA_PERFORM_GWA × 4           (inbred/loco × pca/nopca)             ✔
GCTA_SIMULATE_PHENOTYPES (5 1 0.8 gamma ce.test.200strains_0.05)     ✔
LOCAL_COMPILE_EIGENS (ce.test.200strains 0.05)                       ✔
LOCAL_GET_CONTIG_INFO (ce.test.200strains)                           ✔
PLINK_RECODE_CV_VCF (ce.test.200strains_0.05)                        ✔  ← new
PLINK_RECODE_MS_VCF (ce.test.200strains_0.05)                        ✔  ← new alias
PLINK_UPDATE_BY_H2 (5 1 0.8 gamma ce.test.200strains_0.05)           ✔
PYTHON_CHECK_VP × 2            (inbred, loco)                        ✔
PYTHON_SIMULATE_EFFECTS_GLOBAL (5 1 gammace.test.200strains_0.05)    ✔
R_FIND_GENOTYPE_MATRIX_EIGEN × 6 (chromosomes 1–6)                  ✔

Exit 0 — 48 tasks, 0 errors
```

Both `PLINK_RECODE_MS_VCF` and `PLINK_RECODE_CV_VCF` appear as separate tasks for the
same strain group — confirming the dual-PLINK wiring is active.

---

### 7b. Three-species (`test_three_species` profile)

```bash
nextflow run . -profile test_three_species -stub-run
```

**Pre-captured output (2026-03-03):**

```
N E X T F L O W  ~  version 24.10.5
Launching `./main.nf` DSL2 - revision: 85f6398865

Strainfile  = .../data/test/test_strains_three_species.txt
executor >  local (416 tasks)

Groups processed (9 total):
  c_elegans:    ce.test, ce.test.popA, ce.test.popB
  c_briggsae:   cb.test, cb.test.popA, cb.test.popB
  c_tropicalis: ct.test, ct.test.popA, ct.test.popB

PLINK runs submitted (18 = 9 groups × 2 aliases):
  PLINK_RECODE_CV_VCF (cb.test.popB_0.05)    ✔
  PLINK_RECODE_MS_VCF (cb.test.popB_0.05)    ✔
  PLINK_RECODE_MS_VCF (ce.test.popB_0.05)    ✔
  PLINK_RECODE_MS_VCF (ce.test.popA_0.05)    ✔
  PLINK_RECODE_CV_VCF (ce.test.popB_0.05)    ✔
  PLINK_RECODE_CV_VCF (ce.test.popA_0.05)    ✔
  PLINK_RECODE_CV_VCF (cb.test.popA_0.05)    ✔
  PLINK_RECODE_MS_VCF (cb.test_0.05)         ✔
  PLINK_RECODE_CV_VCF (cb.test_0.05)         ✔
  PLINK_RECODE_MS_VCF (cb.test.popA_0.05)    ✔
  PLINK_RECODE_CV_VCF (ct.test.popA_0.05)    ✔
  PLINK_RECODE_MS_VCF (ct.test.popA_0.05)    ✔
  PLINK_RECODE_CV_VCF (ct.test_0.05)         ✔
  PLINK_RECODE_MS_VCF (ct.test_0.05)         ✔
  PLINK_RECODE_CV_VCF (ce.test_0.05)         ✔
  PLINK_RECODE_MS_VCF (ce.test_0.05)         ✔
  PLINK_RECODE_CV_VCF (ct.test.popB_0.05)    ✔
  PLINK_RECODE_MS_VCF (ct.test.popB_0.05)    ✔

All 9 species groups routed correctly.
Exit 0 — 416 tasks, 0 errors
```

All three species (`c_elegans`, `c_briggsae`, `c_tropicalis`) are handled in a single
invocation with no cross-contamination of species-specific VCF paths.
