---
title: "Verifying Simulation Outputs"
subtitle: "Using bin/verify_db.R to check Parquet database completeness after a run"
date: last-modified
format:
  html:
    embed-resources: true
---

After a pipeline run completes, `bin/verify_db.R` performs a structured
completeness check on the Parquet database under `{outputDir}/db/`. It answers
a single question — *"does the database contain everything the architecture
asked for?"* — and never errors out. It is safe to embed in a Quarto report
chunk alongside the pipeline execution step.

```{mermaid}
%%{init: {'themeVariables': {'fontSize': '16px'}}}%%
flowchart LR
    classDef data fill:#f5f5f5,stroke:#999,color:#333
    classDef tool fill:#4e9af1,stroke:#2c6fad,color:#fff
    classDef ok fill:#48b96e,stroke:#2d8150,color:#fff
    classDef bad fill:#e07b3a,stroke:#a3572a,color:#fff

    DB(["{outputDir}/db/"]):::data
    ARGS(["strainfile + nqtl/h2/effect/reps/cv_*"]):::data
    VD["verify_db()"]:::tool
    R["result list<br/>(summary, deltas, errors)"]:::data
    P["plot_* helpers"]:::tool

    C(["status: complete"]):::ok
    I(["status: incomplete"]):::bad
    E(["status: error"]):::bad
    D(["status: discovery"]):::ok

    DB --> VD
    ARGS --> VD
    VD --> R
    R --> P
    R --> C & I & E & D
```

## Two Modes

**Expectation mode** — you pass the full architecture (`strainfile`, `nqtl`,
`h2`, `effect`, `reps`, `cv_maf`, `cv_ld`). `verify_db()` enumerates every
expected trait and mapping ID using `generate_trait_id()` and
`generate_mapping_id()`, queries the DB for what is actually present, and
reports per-cell deltas and any missing IDs. Final `status` is one of
`"complete"`, `"incomplete"`, or `"error"`.

**Discovery mode** — no architecture args are passed. The grid is derived from
`mappings_metadata.parquet` and only observed distributions are reported. Use
this when you inherit a database and want to know what is inside it. Final
`status` is `"discovery"`.

If only *some* architecture args are passed, `verify_db()` fails closed —
it emits a warning, skips per-ID enumeration for the missing dimensions, and
still returns a result (rather than guessing).

## Non-Failing Semantics

`verify_db()` is designed to be safe inside a Quarto report chunk that runs
unconditionally at the end of a long pipeline invocation. Every internal step
is wrapped so that any error or warning is recorded in `result$errors` rather
than thrown. If the DB does not exist, is partially written, or contains
corrupted Parquet files, the function still returns a result list — the
`status` field indicates `"error"` and the `errors` tibble explains why. Plot
helpers follow the same rule: a failed plot returns a `ggplot` whose title
describes the failure, so `print()` calls never break a rendered document.

## Interactive Usage

From an R session or a `.qmd` chunk:

```{r}
#| eval: false
source("bin/verify_db.R")

result <- verify_db(
  base_dir   = "Sims_ce-cb-ct_nqtl1-50_h2grid_50reps/db",
  strainfile = "data/sims_ce-cb-ct_nqtl1-50_h2grid_50reps/strains_three_species.tsv",
  nqtl       = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 25, 50),
  h2         = c(0.2, 0.4, 0.6, 0.8),
  effect     = "gamma",
  reps       = 25,
  cv_maf     = 0.05,
  cv_ld      = 0.8
)

.vdb_print_summary(result)    # human-readable summary to stdout
print(plot_n_mappings_by_param(result))
print(plot_n_traits_by_param(result))
print(plot_completeness_heatmap(result))
print(plot_marker_set_coverage(result))
```

The same call in discovery mode simply omits the architecture arguments:

```{r}
#| eval: false
source("bin/verify_db.R")
result <- verify_db(base_dir = "Sims_ce-cb-ct_nqtl1-50_h2grid_50reps/db")
```

## CLI Usage

`bin/verify_db.R` is also a standalone script. It returns exit code `0` on
`complete` or `discovery`, `1` on `incomplete`, and `2` on `error`:

```bash
Rscript bin/verify_db.R \
  --base_dir Sims_ce-cb-ct_nqtl1-50_h2grid_50reps/db \
  --strainfile data/sims_ce-cb-ct_nqtl1-50_h2grid_50reps/strains_three_species.tsv \
  --nqtl 1,2,3,4,5,6,7,8,9,10,25,50 \
  --h2 0.2,0.4,0.6,0.8 \
  --effect gamma \
  --reps 25 \
  --cv_maf 0.05 \
  --cv_ld 0.8
```

| Flag | Description |
|------|-------------|
| `--base_dir` | Path to the `{outputDir}/db/` directory (required) |
| `--strainfile` | Strainfile TSV (triggers expectation mode) |
| `--nqtl`, `--h2`, `--effect` | Comma-separated values matching the run parameters |
| `--reps`, `--cv_maf`, `--cv_ld` | Scalar architecture parameters |
| `--max_enumeration_rows` | Safety guard on the expected-grid row count (default 500,000) |
| `--quiet` | Suppress the printed summary |

Discovery mode: omit every flag except `--base_dir`.

When run in expectation mode, the script prints a human-readable summary to stdout and exits with code `0` if the DB is complete, `1` if incomplete, and `2` if an error was encountered. The output should look something like this:

```
2026-04-13 15:17:47 [INFO] Opened database with 3 marker sets, 14400 mapping partitions
verify_db: Sims_ce-cb-ct_nqtl1-50_h2grid_50reps/db
mode: expectation   status: complete

  component                      expected   observed    missing        pct
  marker_sets_markers                   3          3          0     100.0%
  marker_sets_genotypes                 3          3          0     100.0%
  marker_set_metadata                   3          3          0     100.0%
  traits_metadata_files              3600       3600          0     100.0%
  causal_variants_files              3600       3600          0     100.0%
  phenotypes_files                   3600       3600          0     100.0%
  causal_genotypes_files             3600       3600          0     100.0%
  mapping_data_files                14400      14400          0     100.0%
  mapping_meta_files                14400      14400          0     100.0%
  mappings_metadata_rows            14400      14400          0     100.0%

errors: 0   warnings: 0
```

### HPC CLI Usage

On HPC clusters, the intended usage is to call `bin/verify_db.R` from a interactive node with a singularity container. 

Start an interactive session:

```bash
interact -n 1 -c 1 -a eande106 -m 64G -p parallel -t 30
```

Export the singularity cache dir to avoid recreating the container:

```bash
export NXF_SINGULARITY_CACHEDIR=/vast/eande106/singularity
```

Load the singularity module

```bash
module load singularity
```

Set the singularity image path. Nextflow caches images under
`$NXF_SINGULARITY_CACHEDIR` with a mangled filename — verify the exact name
with `ls $NXF_SINGULARITY_CACHEDIR | grep migrate-sims` before setting:

```bash
SIMG=$NXF_SINGULARITY_CACHEDIR/andersenlab-migrate-sims-20260202.img
```

Start an interactive shell inside the container, binding **every** filesystem
root the script needs to reach — the repo, the DB output, and the strainfile.
If your repo lives in `/home/$USER/...`, add `--bind /home/$USER` (the default
singularity binds are not enough on Rockfish):

```bash
singularity shell \
  --bind /vast/eande106 \
  --bind /scratch4/eande106 \
  --bind /home/$USER \
  --pwd $PWD \
  $SIMG
```

Then from inside the container, `cd` to the repo and export
`NEMASCAN_R_LIB` so `bin/verify_db.R` can locate the shared R library. The
script's autodetection (walking up from its own path to find `../R`) can
fail inside the container when the working directory is not the project
root, so setting it explicitly is the most reliable path:

```bash
cd /path/to/nemascan-sims-nf
export NEMASCAN_R_LIB=$(pwd)/R
Rscript bin/verify_db.R --base_dir /path/to/run/db ...
```

## Result Structure

`verify_db()` returns a list with the following slots. Each slot is safe to
access even when the DB is missing or corrupted — missing data is represented
as `NULL` or a zero-row tibble, not a hard error.

| Slot | Type | Content |
|------|------|---------|
| `status` | character | `"complete"`, `"incomplete"`, `"discovery"`, or `"error"` |
| `mode` | character | `"expectation"` or `"discovery"` |
| `base_dir` | character | The path checked |
| `run_at` | POSIXct | Timestamp of the verification run |
| `expected_params` | list | Architecture arguments as received |
| `observed_mappings` | tibble | One row per `(population, nqtl, h2, effect, algorithm, pca)` cell with observed counts |
| `observed_traits` | tibble | One row per unique trait ID present in metadata |
| `mappings_delta` | tibble | Expected vs observed per cell (expectation mode only) |
| `traits_delta` | tibble | Expected vs observed trait counts (expectation mode only) |
| `missing_traits` | tibble | Specific trait IDs expected but not found |
| `missing_mappings` | tibble | Specific mapping IDs expected but not found |
| `summary` | tibble | 10-row file-count summary (see below) |
| `marker_set_coverage` | tibble | Per-population marker set metadata completeness |
| `errors` | tibble | `level` / `step` / `message` for every recorded error or warning |

The `summary` tibble aggregates file counts for the 10 artifacts on disk:
`marker_sets_markers`, `marker_sets_genotypes`, `marker_set_metadata`,
`traits_metadata_files`, `causal_variants_files`, `phenotypes_files`,
`causal_genotypes_files`, `mapping_data_files`, `mapping_meta_files`, and
`mappings_metadata_rows`. Each row has `expected`, `observed`, `missing`, and
`pct_complete` columns — so a quick glance at `result$summary` reveals the
component that is lagging.

## Plot Helpers

Four plot functions accept a `verify_db()` result and return a `ggplot`
object. All four use `facet_grid(algorithm + pca ~ population)` to cover the
full cartesian parameter grid at once:

| Function | Purpose |
|---------|---------|
| `plot_n_mappings_by_param(result)` | Bar chart of observed mappings per `nqtl × h2`, faceted by population and mapping mode |
| `plot_n_traits_by_param(result)` | Same layout for distinct trait counts |
| `plot_completeness_heatmap(result)` | Heatmap of `pct_complete` per architecture cell (expectation mode only) |
| `plot_marker_set_coverage(result)` | Bar chart of marker set metadata file counts per population |

Every helper runs inside an internal safety wrapper — a failed plot returns
a placeholder `ggplot` whose title describes the failure, so unconditional
`print()` calls never break a rendered Quarto document.

## Integration with Quarto Reports

The intended placement is the final "verify outputs" phase of a production
run report. Source the script once, call `verify_db()` with the same
parameters the pipeline was invoked with, print the summary, and render the
plots:

```{r}
#| eval: false
#| echo: false
source("bin/verify_db.R")
```

```{r}
#| eval: false
result <- verify_db(
  base_dir   = file.path(params$output_dir, "db"),
  strainfile = file.path(params$data_dir, "strains_three_species.tsv"),
  nqtl       = scan(file.path(params$data_dir, "nqtl.csv")),
  h2         = scan(file.path(params$data_dir, "h2.csv")),
  effect     = readLines(file.path(params$data_dir, "effect_sizes.csv")),
  reps       = params$reps,
  cv_maf     = params$cv_maf,
  cv_ld      = params$cv_ld
)
.vdb_print_summary(result)
```

```{r}
#| eval: false
#| fig-width: 10
#| fig-height: 7
print(plot_n_mappings_by_param(result))
print(plot_n_traits_by_param(result))
print(plot_completeness_heatmap(result))
print(plot_marker_set_coverage(result))
```

Because `verify_db()` never errors out, these chunks can be unconditionally
rendered — the `status` field and `errors` tibble carry any bad news in a
structured form rather than breaking the render.

## Unit Tests

`tests/testthat/test-verify_db.R` exercises the script against a tempdir
fixture DB built from the real `database.R` hash helpers. It does **not**
require `TEST_DB_DIR` or any pipeline output — it runs alongside the other
unit tests as part of `Rscript tests/run_tests.R`. The 8 tests cover:

| Test | Asserts |
|------|---------|
| Empty DB (discovery) | Returns a well-formed result with `status = "discovery"` |
| Complete fixture | `status = "complete"`; all summary components at 100 % |
| Incomplete fixture | Drops 4 mappings; `status = "incomplete"`; `missing_mappings` lists them |
| Discovery grid derivation | Observed grid matches the fixture without architecture args |
| Corrupted Parquet | Zero-byte `mappings_metadata.parquet` recorded as an error, not thrown |
| `max_enumeration_rows` guard | Huge architecture request surfaces a guard warning |
| Partial architecture args | Fails closed with a warning, returns a result |
| Plot helpers | All four helpers return `ggplot` objects on both good and bad input |

::: {.callout-tip}
## See also
- [Output Validation](testing-output-validation.qmd) — cross-validation tests,
  `renv.lock`, and GCTA thread-count determinism.
- [Database Structure](database-structure.qmd) — the Parquet layout that
  `verify_db()` iterates over.
:::
